---
title: "NIRS Statistical Analysis - Complete Pipeline"
author: "Data Analysis Project"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, fig.height = 6)
```

# NIRS Statistical Analysis Pipeline

## 1. Load Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)      # Data manipulation and visualization
library(lme4)           # Linear Mixed-Effects Models (LMM)
library(lmerTest)       # Provides p-values and ANOVA for LMM
library(emmeans)        # Post-hoc comparisons and marginal means
library(performance)    # Model diagnostics
library(ggplot2)        # Advanced visualization
```

---

## 2. Configuration

```{r configuration}
# Set the path to your file
# file_path <- "C:/Program Files/DigiMove/DigiMove/DataAnalysisProject/Analysis_Results/hb_summary_final_semaxone.csv"

# Choose your variable of interest
target_variable <- "EnvelopeMean_HbR_uM"

# For demonstration, we'll create synthetic data matching your study design
# Replace this section with: df <- read_csv2(file_path) when using real data
```

---

## 3. Load and Prepare Data

```{r data-loading}
# Create synthetic data matching the experimental design and statistics from Table 1
set.seed(42)

# Function to generate data for a condition
generate_condition_data <- function(n_subjects, n_trials, mean_val, sd_val, mode, intensity) {
  expand.grid(
    Subject = paste0("Sub", sprintf("%02d", 1:n_subjects)),
    Trial = 1:n_trials
  ) %>%
    mutate(
      Contraction_Mode = mode,
      Intensity = intensity,
      EnvelopeMean_HbR_uM = rnorm(n_subjects * n_trials, mean = mean_val, sd = sd_val)
    )
}

# Combine all conditions
df_raw <- bind_rows(
  generate_condition_data(23, 3, 2.645, 1.674, "Concentric", "30%"),
  generate_condition_data(23, 3, 3.492, 1.900, "Concentric", "50%"),
  generate_condition_data(23, 3, 2.669, 1.897, "Eccentric", "30%"),
  generate_condition_data(23, 3, 3.001, 1.893, "Eccentric", "50%")
)

# Data processing pipeline
df_clean <- df_raw %>%
  mutate(
    NIRS_Value = .data[[target_variable]],
    Subject = as.factor(Subject),
    Contraction_Mode = as.factor(Contraction_Mode),
    Intensity = as.factor(gsub("%", "", Intensity))
  ) %>%
  drop_na(NIRS_Value)

# Display first few rows
cat("Data structure:\n")
head(df_clean, 10)
cat("\nSample sizes per condition:\n")
table(df_clean$Contraction_Mode, df_clean$Intensity)
```

---

## 4. Descriptive Statistics

```{r descriptive-stats}
# Computing descriptive statistics
stats_summary <- df_clean %>%
  group_by(Contraction_Mode, Intensity) %>%
  summarise(
    Mean = mean(NIRS_Value),
    SD = sd(NIRS_Value),
    SEM = sd(NIRS_Value) / sqrt(n()),
    N = n(),
    CI_lower = Mean - 1.96 * SEM,
    CI_upper = Mean + 1.96 * SEM,
    .groups = 'drop'
  )

cat("\n=== DESCRIPTIVE STATISTICS (Table 1) ===\n")
print(stats_summary)

# Save as formatted table
stats_table <- stats_summary %>%
  select(Contraction_Mode, Intensity, Mean, SD, N) %>%
  mutate(across(Mean:SD, ~round(., 3)))
knitr::kable(stats_table, caption = "Table 1: Descriptive Statistics for NIRS Analysis")
```

---

## 5. Visualization - Boxplot with Individual Points

```{r boxplot}
ggplot(df_clean, aes(x = Intensity, y = NIRS_Value, fill = Contraction_Mode)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), 
              size = 1, alpha = 0.3) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    legend.position = "top"
  ) +
  labs(
    title = paste("NIRS Analysis:", target_variable),
    subtitle = "Comparison by Contraction Mode and Intensity",
    x = "Intensity (% MVC)",
    y = paste(target_variable, "(µM)"),
    fill = "Contraction Mode"
  ) +
  scale_fill_manual(values = c("Concentric" = "#4e79a7", "Eccentric" = "#f28e2b"))
```

---

## 6. Linear Mixed-Effects Model Fitting

```{r model-fitting}
# Fit the Linear Mixed-Effects Model
# Fixed effects: Contraction_Mode, Intensity, and their interaction
# Random effects: Random intercept for each subject (accounts for repeated measures)
model <- lmer(NIRS_Value ~ Contraction_Mode * Intensity + (1|Subject), data = df_clean)

cat("\n=== MODEL SUMMARY ===\n")
print(summary(model))
```

---

## 7. ANOVA (Type III Wald F-tests)

```{r anova-results}
cat("\n\n=== ANOVA RESULTS (Type III Wald F-tests) ===\n")
anova_table <- anova(model)
print(anova_table)

# Extract and interpret results - Handle both older and newer lmerTest formats
cat("\n--- INTERPRETATION ---\n")

# Convert to data frame for easier handling
anova_df <- as.data.frame(anova_table)

# Main Effect: Contraction_Mode
if("Contraction_Mode" %in% rownames(anova_df)) {
  cm_result <- anova_df["Contraction_Mode", ]
  denom_df <- if("Denom. DF" %in% names(cm_result)) as.numeric(cm_result$`Denom. DF`) else NA
  f_val <- as.numeric(cm_result$`F value`)
  p_val <- as.numeric(cm_result$`Pr(>F)`)
  
  cat("Main Effect: Contraction_Mode\n")
  if(!is.na(denom_df)) {
    cat(sprintf("F(1, %d) = %.3f, p = %.4f\n", as.integer(denom_df), f_val, p_val))
  } else {
    cat(sprintf("F = %.3f, p = %.4f\n", f_val, p_val))
  }
}

# Main Effect: Intensity
if("Intensity" %in% rownames(anova_df)) {
  cat("\nMain Effect: Intensity\n")
  int_result <- anova_df["Intensity", ]
  denom_df <- if("Denom. DF" %in% names(int_result)) as.numeric(int_result$`Denom. DF`) else NA
  f_val <- as.numeric(int_result$`F value`)
  p_val <- as.numeric(int_result$`Pr(>F)`)
  
  if(!is.na(denom_df)) {
    cat(sprintf("F(1, %d) = %.3f, p = %.4f\n", as.integer(denom_df), f_val, p_val))
  } else {
    cat(sprintf("F = %.3f, p = %.4f\n", f_val, p_val))
  }
}

# Interaction: Contraction_Mode × Intensity
if("Contraction_Mode:Intensity" %in% rownames(anova_df)) {
  cat("\nInteraction: Contraction_Mode × Intensity\n")
  interaction_result <- anova_df["Contraction_Mode:Intensity", ]
  denom_df <- if("Denom. DF" %in% names(interaction_result)) as.numeric(interaction_result$`Denom. DF`) else NA
  f_val <- as.numeric(interaction_result$`F value`)
  p_val <- as.numeric(interaction_result$`Pr(>F)`)
  
  if(!is.na(denom_df)) {
    cat(sprintf("F(1, %d) = %.3f, p = %.4f\n", as.integer(denom_df), f_val, p_val))
  } else {
    cat(sprintf("F = %.3f, p = %.4f\n", f_val, p_val))
  }
}
```

---

## 8. Post-hoc Pairwise Comparisons (Bonferroni Correction)

```{r posthoc-comparisons}
cat("\n\n=== PAIRWISE COMPARISONS (Bonferroni Adjustment) ===\n")

# Compare Concentric vs Eccentric for each Intensity level
posthoc_results <- emmeans(model, pairwise ~ Contraction_Mode | Intensity, 
                           adjust = "bonferroni")

cat("\n--- Estimated Marginal Means ---\n")
print(posthoc_results$emmeans)

cat("\n--- Contrasts (Concentric - Eccentric) ---\n")
print(posthoc_results$contrasts)

# Extract specific contrasts for reporting - Handle different emmeans versions
cat("\n--- INTERPRETATION ---\n")

# Convert to data frame safely
contrasts_df <- as.data.frame(posthoc_results$contrasts)

# Check which column names exist (varies by emmeans version)
contrast_col <- if("contrast" %in% names(contrasts_df)) "contrast" else 
                if("Comparison" %in% names(contrasts_df)) "Comparison" else NULL

estimate_col <- if("estimate" %in% names(contrasts_df)) "estimate" else 
                if("Estimate" %in% names(contrasts_df)) "Estimate" else NULL

intensity_col <- if("Intensity" %in% names(contrasts_df)) "Intensity" else 
                 if("intensity" %in% names(contrasts_df)) "intensity" else NULL

pvalue_col <- if("p.value" %in% names(contrasts_df)) "p.value" else 
              if("Pr(>|t|)" %in% names(contrasts_df)) "Pr(>|t|)" else 
              if("Pr(>F)" %in% names(contrasts_df)) "Pr(>F)" else NULL

# CI columns - check multiple variations
ci_lower <- if("lower.CL" %in% names(contrasts_df)) "lower.CL" else 
            if("asymp.LCL" %in% names(contrasts_df)) "asymp.LCL" else 
            if("Lower.CL" %in% names(contrasts_df)) "Lower.CL" else NULL

ci_upper <- if("upper.CL" %in% names(contrasts_df)) "upper.CL" else 
            if("asymp.UCL" %in% names(contrasts_df)) "asymp.UCL" else 
            if("Upper.CL" %in% names(contrasts_df)) "Upper.CL" else NULL

# Try to extract and print results
if(!is.null(estimate_col) && !is.null(pvalue_col) && !is.null(intensity_col)) {
  for(i in 1:nrow(contrasts_df)) {
    intensity <- contrasts_df[[i, intensity_col]]
    estimate <- contrasts_df[[i, estimate_col]]
    p_val <- contrasts_df[[i, pvalue_col]]
    
    # Try to get CI bounds if available
    if(!is.null(ci_lower) && !is.null(ci_upper)) {
      ci_l <- contrasts_df[[i, ci_lower]]
      ci_u <- contrasts_df[[i, ci_upper]]
      cat(sprintf("Intensity %s: Diff = %.3f µM, 95%% CI [%.3f, %.3f], p = %.4f\n",
                  intensity, estimate, ci_l, ci_u, p_val))
    } else {
      cat(sprintf("Intensity %s: Diff = %.3f µM, p = %.4f\n",
                  intensity, estimate, p_val))
    }
  }
} else {
  # Fallback: just print the data frame
  cat("Contrasts table:\n")
  print(contrasts_df)
}
```

---

## 9. Estimated Marginal Means for Visualization

```{r emmeans-for-plot}
# Get estimated marginal means
emmeans_data <- emmeans(model, ~ Contraction_Mode * Intensity) %>%
  as.data.frame() %>%
  mutate(
    CI_lower = emmean - 1.96 * SE,
    CI_upper = emmean + 1.96 * SE
  )

cat("\n=== ESTIMATED MARGINAL MEANS (for plotting) ===\n")
print(emmeans_data)
```

---

## 10. INTERACTION PLOT (Main Visualization)

```{r interaction-plot}
# Create interaction plot with error bars
interaction_plot <- ggplot(emmeans_data, 
                           aes(x = Intensity, y = emmean, 
                               color = Contraction_Mode, 
                               group = Contraction_Mode)) +
  # Add lines connecting points
  geom_line(size = 1.2) +
  # Add points for each mean
  geom_point(size = 4) +
  # Add error bars (95% CI)
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper), 
                width = 0.1, size = 1) +
  # Customize theme
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 11),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "lightgray", size = 0.3),
    panel.grid.minor = element_blank()
  ) +
  # Labels and scales
  labs(
    title = "Interaction Effect on HbR Concentration",
    subtitle = "Estimated Marginal Means with 95% Confidence Intervals",
    x = "Intensity (% MVC)",
    y = "Deoxygenated Hemoglobin (HbR) [µM]",
    color = "Contraction Mode"
  ) +
  scale_color_manual(values = c("Concentric" = "#4e79a7", "Eccentric" = "#f28e2b")) +
  # Add minor adjustments for readability
  ylim(min(emmeans_data$CI_lower) - 0.5, max(emmeans_data$CI_upper) + 0.5) +
  scale_x_discrete(labels = c("30" = "30%", "50" = "50%"))

print(interaction_plot)

# Save the plot
ggsave("interaction_plot_HbR.png", plot = interaction_plot, 
       width = 10, height = 6, dpi = 300)
cat("\nInteraction plot saved as 'interaction_plot_HbR.png'\n")
```

---

## 11. Model Diagnostics

```{r model-diagnostics, fig.height=8, fig.width=10}
cat("\n=== MODEL DIAGNOSTICS ===\n")

# Check model assumptions
check_model(model)

# Additional diagnostic plots
par(mfrow = c(2, 2))
plot(model, main = "Residuals vs Fitted")
qqnorm(residuals(model), main = "Q-Q Plot")
qqline(residuals(model))
hist(residuals(model), main = "Distribution of Residuals", xlab = "Residuals")
plot(residuals(model), main = "Residuals vs Order")
par(mfrow = c(1, 1))
```

---

## 12. Effect Sizes and Model Quality

```{r effect-sizes}
cat("\n=== EFFECT SIZES & MODEL QUALITY ===\n")

# Calculate R-squared values
cat("\nModel Fit:\n")
print(performance::model_performance(model))

# Variance explained
cat("\nRandom Effects Variance:\n")
print(VarCorr(model))
```

---

## 13. Summary Report

```{r summary-report}
cat("\n\n========================================================\n")
cat("           COMPREHENSIVE STATISTICAL SUMMARY\n")
cat("========================================================\n\n")

cat("STUDY DESIGN:\n")
cat(sprintf("- Contraction modes: Concentric, Eccentric\n"))
cat(sprintf("- Intensities: 30%%, 50%% of Maximum Voluntary Contraction\n"))
cat(sprintf("- Sample size: N = 23 subjects\n"))
cat(sprintf("- Total observations: N = %d\n\n", nrow(df_clean)))

cat("DESCRIPTIVE STATISTICS:\n")
print(stats_summary %>% select(Contraction_Mode, Intensity, Mean, SD, N))

cat("\n\nANOVA RESULTS:\n")
cat("The Type III ANOVA revealed:\n")

# Safe extraction of p-values
anova_df <- as.data.frame(anova(model))

check_sig <- function(pval, name) {
  if(!is.na(pval) && pval < 0.05) {
    cat(sprintf("✓ Significant main effect of %s (p = %.4f)\n", name, pval))
  } else {
    cat(sprintf("✗ No significant main effect of %s (p = %.4f)\n", name, pval))
  }
}

if("Contraction_Mode" %in% rownames(anova_df)) {
  p_cm <- as.numeric(anova_df["Contraction_Mode", "Pr(>F)"])
  check_sig(p_cm, "Contraction Mode")
}

if("Intensity" %in% rownames(anova_df)) {
  p_int <- as.numeric(anova_df["Intensity", "Pr(>F)"])
  check_sig(p_int, "Intensity")
}

if("Contraction_Mode:Intensity" %in% rownames(anova_df)) {
  p_inter <- as.numeric(anova_df["Contraction_Mode:Intensity", "Pr(>F)"])
  check_sig(p_inter, "Contraction Mode × Intensity Interaction")
}

cat("\n\nPOST-HOC FINDINGS:\n")
contrasts_df <- as.data.frame(posthoc_results$contrasts)

# Safely select columns - handle different emmeans versions
cols_to_select <- c()
if("contrast" %in% names(contrasts_df)) cols_to_select <- c(cols_to_select, "contrast")
if("Intensity" %in% names(contrasts_df)) cols_to_select <- c(cols_to_select, "Intensity")
if("estimate" %in% names(contrasts_df)) cols_to_select <- c(cols_to_select, "estimate")
if("lower.CL" %in% names(contrasts_df)) cols_to_select <- c(cols_to_select, "lower.CL")
if("upper.CL" %in% names(contrasts_df)) cols_to_select <- c(cols_to_select, "upper.CL")
if("p.value" %in% names(contrasts_df)) cols_to_select <- c(cols_to_select, "p.value")

if(length(cols_to_select) > 0) {
  print(contrasts_df %>% 
    select(all_of(cols_to_select)) %>%
    mutate(across(where(is.numeric), ~round(., 4))))
} else {
  print(contrasts_df)
}

cat("\n========================================================\n")
cat("For detailed interpretation, see Section 4.2 of the report.\n")
cat("========================================================\n")
```

---

## Notes

- **Model Formula**: `NIRS_Value ~ Contraction_Mode * Intensity + (1|Subject)`
- **Fixed Effects**: Contraction mode (2 levels), Intensity (2 levels), and their interaction
- **Random Effects**: Random intercept for each subject (accounts for repeated measures)
- **Multiple Comparisons Correction**: Bonferroni adjustment applied to post-hoc tests
- **Data Source**: EnvelopeMean_HbR_uM from NIRS signal processing pipeline
